Facility and Time Step Discretization Effects in Fuel Cycle Simulation

### 2015-10-02 paper.tex
edits from dad up to scenario description section

Due to the diversity of fuel cycle simulators' modeling assumptions direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions involving reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of a few different modeling
choices that may sometimes be taken for granted in the fuel cycle modeling
domain.  Four scenarios are compared using combinations of either fleet-based
or individual reactors with either monthly or quarterly (3-month) time steps.
The Cyclus fuel cycle simulator is used as a level playing field to compare
the scenarios.  When under supply-constraint pressure, the four cases exhibit
noticably different behavior.  Otherwise, they generally behaved very
similarly.  Fleet-based modeling is more efficient in supply-constrained
environments at the expense of losing insight on issues such as realistically
suboptimal fuel distribution and challenges in reactor refueling cycle
staggering.  Finer-grained time steps also enable more efficient material use
in supply-constrained environments.  Large simulations with fleet-based
reactors run much more quickly than their individual reactor counterparts.

@@

Due to the diversity of fuel cycle simulators' modeling assumptions direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different due to inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of a few different modeling choices that may sometimes be
taken for granted in the fuel cycle modeling domain: facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics can
allow for making more deliberate decisions with respect to tradeoffs such as
performance vs. realism.

@@

Traditionally, many fuel cycle simulators have used a system dynamics approach
\cite{forrester_industrial_1961} to modeling. These simulators model several
"stocks" representing isotopes or isotope groups that have levels that change
over time due to "flows" between them.  The flows are determined by equations
that are a function of the state of the system (i.e. the system's stocks).  A
constructed system is then solved in discrete or continuous time depending on
the limitations of the software and constraints imposed by the model.  In
system dynamics simulators, the number of stocks and potential flows between
them is generally a static property of the modeled system, making it somewhat
difficult to swap different reactor models in and out of the same scenario
without wiring them into the simulator manually.

System dynamics based fuel cycle simulators generally use a collection of
stocks to represent the state of groups of like-facilities commonly referred
to as "fleets".  As the simulation steps through time, the levels of stocks
(e.g. reactor fresh fuel inventory, repository waste inventories, etc.) are
adjusted according to the calculated flow values.  Simulators that model
facility fleets built on a system dynamics foundation include VISION
\cite{jacobson_verifiable_2010}, DANESS \cite{van_den_durpel_daness_2009},
DYMOND (predecessor to DANESS), and ORION \cite{gregg_orion_2012} among
others.  In fleet-based modeling:

@@

This work investigates in detail some differences between fleet vs. individual
reactor modeling.  The difference between these two modeling styles in terms
of their affects on output of fuel cycle simulators in general is currently
unknown. This analysis is designed to provide both quantitative and
qualitative insight about how fleet-based continuous flow models differ from
individual facility discrete flow models.

Many fuel cycle simulators step through time in discrete steps.  The duration
of a single time step can have significant impact on the dynamics of a
simulation.  The time step duration can affect the magnitude of individual
material transfers between facilities and correspondingly affects the noise
level in facility material inventories and flows.  It can also act as a
minimum bound for outage durations. The effects of the time step duration are
also investigated together with fleet vs facility modeling.

@@

One of the major motivations for the development of the Cyclus fuel cycle
simulator \cite{cyclus_2015} was creating a simulation environment that is
more expressive than more traditional fuel cycle simulators.  Cyclus was
designed with a plugin-style architecture that allows different models for
facility types, known as \emph{archetypes} to be swapped in and out trivially.

As a Cyclus simulation walks through each time step, facilities engage in a
market-like dynamic resource exchange (DRE).  They broadcast requests for
material and suppliers respond with bids. Requesters then express preferences
between all their bids, and the DRE then creates a network flow problem that
is optimized to resolve which transactions actually occur.  The plugin
architecture combined with the DRE creates a powerful and flexible environment
for the natural comparison of different modeling paradigms.  Cyclus'
agent-based architecture in concert with its DRE has the ability to model
fleet-based facilities.  Additionally, however, Cyclus can model facilities
individually with discrete flows enabling the investigation of interesting
real-world effects such as competition, reactor cycle staggering, and
individual reactor outages.

Cyclus was designed to operate natively in unix/linux type environments with
flexible and scriptable usage and input file format.  Cyclus stores its
diverse generated data in a powerful open-source database format.  In addition
to information about material flows and facility deployments, Cyclus'
single-file database contains the simulation input file, version information
about the archetypes used in a simulation, the version of Cyclus used to run
the simulation, and the installed versions of Cyclus dependencies. This and
other features of Cyclus lend themselves well to automation, research
provenance and reproducibility, and large-scale computing.

@@

Because the standard package of archetypes included with Cyclus does not
include a fleet-based reactor model, one was necessarily developed. A
simulation scenario was created that is roughly based on prior and ongoing
work in the Department of Energy (DOE) fuel cycle options campaign.  A few
phenomena related to the different modeling choices in each of the four cases
are identified and used as a basis for comparison.  Each of these things is
described in more detail in the following sections.

@@

A fleet reactor facility was created for Cyclus.  The flexibility of Cyclus
makes it relatively straight forward to create plug-in archetypes that
live anywhere along the spectrum from individual, discrete facilities to
continuous flow fleet facilities.  In order to contrast well with individual,
single-facility granularity modeling, the fleet model created here falls more
toward the aggregate, continuous flow side of the spectrum.

@@

        $D$ is the discharge rate (kg per time step), $B$ is the batch size
        (kg) for a single reactor's batch, $L$ is the cycle length (including
        refueling time) in time steps, $N_{r}$ is the number of reactors in
        the fleet, $C_{inv}$ is the total amount of fuel in all reactor cores
        (kg), and $C_{cap}$ is the total fuel capacity for all reactor cores
        (kg).  The last fractional term accounts for any amount of
        underfueling that may be occuring resulting in lower fuel burning
        rates due to unfueled reactors.

@@

    \item When a reactor in the fleet is retired, a full core of material is
        discharged - even if there is some fraction the fleet with unfueled
        cores.

@@

One characteristic of this fleet reactor implementation is that it has perfect
fuel sharing and perfect cycle staggering among the individual reactors it
models.  The first discharge of spent fuel from fleets begins immediately on
the first time step of operation rather than after a complete refueling cycle.
Fuel shortages also only cause proportional loss of capacity rather than
whole-reactor quantized shutdown.  The full implementation of the fleet-based
reactor is publicy available for download and use \cite{Carlsen2015}.

@@

A pre-release of Cyclus version 1.4 was used for this analysis.  Archetypes
used were the standard ones provided with Cyclus in the included Cycamore
library - also a pre-release of version 1.4.  A custom storage facility was
used because Cycamore did not yet provide one.  The standard cycamore reactor
was used for both LWR and SRF reactor types in cases MI and QI.  Correspondingly
the custom fleet-based reactor described earlier was used for both reactor
types in cases MF and QF.  Additional configuration for the reactor types is
shown in Table \ref{tab:reactors}.

@@

There are various low level phenomena that have potential to affect the larger
outcome of simulations with respect to facility and time step modeling
decisions. Several effects are described below that arise from these
different modeling choices.

>>>>

Due to the diversity of fuel cycle simulator modeling assumptions, direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions involving reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of selected modeling choices
that may sometimes be taken for granted in the fuel cycle modeling domain.
Four scenarios are compared using combinations of either fleet-based or
individual reactors with either monthly or quarterly (3-month) time steps.
The Cyclus fuel cycle simulator's plug-in facility cability along with its
market-like dynamic material routing allow it to be used as a level playing
field for comparing the scenarios.  When under supply-constraint pressure, the
four cases exhibit noticably different behavior.  Fleet-based modeling is more
efficient in supply-constrained environments at the expense of losing insight
on issues such as realistically suboptimal fuel distribution and challenges in
reactor refueling cycle staggering.  Finer-grained time steps also enable more
efficient material use in supply-constrained environments.  Large simulations
with fleet-based reactors run much more quickly than their individual reactor
counterparts.

@@

The diversity of assumptions embedded within various fuel cycle simulators pose challenges for direct
comparison and benchmarking.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different due to inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of selected modeling choices relative to facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics can
allow for making more deliberate decisions with respect to tradeoffs such as
computational investment vs. realism.

@@

Traditionally, many fuel cycle simulators have used a system dynamics approach
\cite{forrester_industrial_1961} to modeling. These simulators model several
"stocks" representing isotopes or isotope groups that have levels that change
over time due to "flows" between them.  The flows are determined by equations
that are a function of the state of the system (i.e. the system's stocks).  A
constructed system is then solved in discrete or continuous time depending on
the limitations of the software and constraints imposed by the model.  In
system dynamics based simulators, the number of stocks and potential flows between
them is generally a static property of the modeled system, making it somewhat
difficult to swap different reactor models in and out of the same scenario
without wiring them into the simulator manually.

System dynamics based fuel cycle simulators generally use a collection of
stocks to represent the state of groups of like-facilities commonly referred
to as "fleets".  As the simulation steps through time, the levels of stocks
(e.g. reactor fresh fuel inventory, repository waste inventories, etc.) are
adjusted according to the calculated flow values.  Fuel cycle simulators that model
facility fleets built on a system dynamics foundation include VISION
\cite{jacobson_verifiable_2010}, DANESS \cite{van_den_durpel_daness_2009},
DYMOND, and ORION \cite{gregg_orion_2012} among
others.  In fleet-based modeling:

@@

This work investigates in detail some differences between fleet vs. individual
reactor modeling.  The difference between these two modeling styles in terms
of their affects on output of fuel cycle simulators in general is not well
known. This analysis is designed to provide both quantitative and
qualitative insight about how fleet-based continuous flow models differ from
individual facility discrete flow models.

Many fuel cycle simulators step through time in discrete steps.  The duration
of a single time step can have significant impact on the dynamics of a
simulation.  The time step duration can affect the magnitude of individual
material transfers between facilities and correspondingly affects the noise
level in facility material inventories and flows.  It can also act as a
minimum bound for outage durations and is a minimum bound in general for state
changes within the entire system. The effects of the time step duration are
also investigated together with fleet vs facility modeling.

@@

One of the major motivations for the development of the Cyclus fuel cycle
simulator \cite{cyclus_2015} was creating a simulation environment that is
more expressive than more traditional fuel cycle simulators.  Cyclus was
designed with a plugin-style architecture that allows different models for
facility types, known as \emph{archetypes}, to be swapped in and out easily.

As a Cyclus simulation walks through each time step, facilities engage in a
market-like dynamic resource exchange (DRE).  They broadcast requests for
material and potential supplying facilities respond with bids. The requesting facilities then
express preferences between all their bids, and the DRE then generates a network
flow problem that is optimized to resolve which transactions actually occur.
The plugin architecture combined with the DRE creates a powerful and flexible
environment for the natural comparison of different modeling paradigms.
Cyclus' module plug-in based architecture in concert with its DRE has the ability to
model fleet-based facilities.  Additionally, however, Cyclus can model
facilities individually with discrete flows enabling the investigation of
interesting real-world effects such as competition, reactor cycle staggering,
and individual reactor outages.

Cyclus was designed to operate natively in unix/linux type environments with
flexible and scriptable usage and input file format.  Cyclus stores its
comprehensive output data in a powerful open-source database format.  In addition
to information about material flows and facility deployments, Cyclus'
single-file database contains the simulation input file, version information
about the archetypes used in a simulation, the version of Cyclus used to run
the simulation, and the installed versions of Cyclus dependencies. This and
other features of Cyclus lend themselves well to automation, research
provenance and reproducibility, and large-scale computing.

@@

Because the standard package of archetypes included with Cyclus does not
include a fleet-based reactor model, one was necessarily developed. A
simulation scenario was created that is roughly based on prior and ongoing
work in the Department of Energy (DOE) fuel cycle options campaign.  A few
phenomena related to the different modeling choices in each of the four cases
are identified and used as a basis for comparison.  These are
described in more detail in the following sections.

@@

A fleet reactor archetype was created for Cyclus.  The flexibility of Cyclus
makes it relatively straight forward to create plug-in archetypes to model
anywhere along the spectrum from individual, discrete facilities to
continuous flow fleet facilities.  In order to contrast well with individual,
single-facility granularity modeling, the fleet archetype created here falls more
toward the aggregate, continuous flow side of the spectrum.

@@

        $D$ is the discharge rate (kg per time step), $B$ is the batch size
        (kg) for a single reactor's batch, $L$ is the cycle length (including
        refueling time) in time steps, $N_{r}$ is the number of reactors in
        the fleet, $C_{inv}$ is the total amount of fuel in all reactor cores
        (kg), and $C_{cap}$ is the total fuel capacity for all reactor cores
        (kg).  The last fractional term accounts for lower fuel burning rates
        due to unfueled reactors.

@@

    \item When a reactor in the fleet is retired, a full core of material is
        discharged - even if there is some fraction of the fleet with unfueled
        cores.

@@

One characteristic of this fleet reactor implementation is that it has perfect
fuel sharing and perfect cycle staggering among the individual reactors it
models.  The first discharge of spent fuel begins immediately on
the first time step of operation rather than after a complete refueling cycle.
Fuel shortages also only cause proportional loss of capacity rather than
whole-reactor quantized shutdown.  The full implementation of the fleet-based
reactor is publicy available for download and use \cite{Carlsen2015}.

@@

A pre-release of Cyclus version 1.4 was used for this analysis.  Archetypes
used were the standard ones provided with Cyclus in the included Cycamore
library - also a pre-release of version 1.4.  A custom storage archetype was
used because Cycamore did not yet provide one.  The standard cycamore reactor
was used for both LWR and SRF reactor types in cases MI and QI.  Correspondingly
the custom fleet-based reactor described earlier was used for both reactor
types in cases MF and QF.  Additional configuration for the reactor types is
shown in Table \ref{tab:reactors}.

@@

There are various low level phenomena that have potential to affect the larger
outcome of simulations with respect to facility discretization and time step modeling
decisions. Several effects are described below that arise from these
different modeling choices.
